\begin{thebibliography}{10}
% (1)
\bibitem{RTCA:DO-178}
RTCA DO-178. 
\newblock Software Considerations in Airborne Systems and Equipment 
Certification. RTCA and EUROCAE, 2011.
%
\bibitem{Hi-Lite:URL}
AdaCore {Hi-Lite} project. 
\newblock \url{http://www.open-do.org/projects/hi-lite/}
% (2)
\bibitem{RTCA:DO-333} 
RTCA DO-333. 
\newblock Formal Methods Supplement to DO-178C and DO-278A. 
RTCA and EUROCAE, 2011.
% (3)
\bibitem{Ada:URL} 
{Ada} reference manual.
\newblock \url{http://www.ada-auth.org/standards/ada12.html}
% (4)
\bibitem{Gnatprove:URL} 
AdaCore {Gnatprove} tool.
\newblock \url{http://www.open-do.org/projects/hi-lite/gnatprove/}
%
\bibitem{Leroy:09}
X.~Leroy. 
Formal Verification of a Realistic Compiler. 
\newblock In {\em Communications of the ACM}, 52(7):107--115, 2009.
%\bibitem{CompCert:URL}
%CompCert certified compiler.
%\newblock \url{http://compcert.inria.fr/}
% (5)
\bibitem{Sireum:URL} 
{Sireum} software analysis platform.
\newblock \url{http://www.sireum.org}
% (6)
\bibitem{Jago:URL} 
{Jago} translation tool.
\newblock \url{https://github.com/sireum/bakar/tree/master/sireum-bakar-jago}
%
\bibitem{Formalization:URL}
Coq code for SPARK 2014 language subset formalization.
\newblock \url{https://github.com/sireum/bakar/tree/master/sireum-bakar-formalization}
% (7)
\bibitem{Marsh:Book94} 
W.~March.
\newblock {\em Formal Semantics of SPARK - Static Semantics}. 
October, 1994.
% (8)
\bibitem{Neil:Book94} 
I.~O'Neill.
\newblock {\em Formal Semantics of SPARK - Dynamic Semantics}.
October, 1994.

\end{thebibliography}
